---
title: "methylTFRAnnotationHg38"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{methylTFRAnnotationHg38}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`methylTFRAnnotationHg38` provides the genome annotations `methylTFR` needs to
compute bias-corrected transcription factor deviation scores on hg38:

- **binding sites**, one `GRanges` per motif, each range extended
    so that methylation can be read across the footprint window;
- **motif GC frequency tables**, recording how each motif's binding
    sites distribute across genome-wide GC quintiles;
- **the genome-wide GC distribution**, which assigns each
    methylation call to a GC bin.

The GC tables are what make the deviation score bias-corrected. A
motif whose binding sites sit in GC-rich sequence will overlap CpG
islands more often than one that does not, and CpG islands are
hypomethylated; without the correction that difference in sequence
composition would read as a difference in factor activity.

The data are too large to ship inside the package, so they are
hosted on AnnotationHub and downloaded on first use. Subsequent
calls are served from the local AnnotationHub cache.

```{r dummy, include = FALSE}
# HIDDEN CHUNK: Set up mock data so the vignette chunks can evaluate
mock_dir <- tempdir()

# 1. Mock TF bindsites
mock_grl <- GenomicRanges::GRangesList(
    mock_motif_1 = GenomicRanges::GRanges("chr1:100-200")
)
saveRDS(mock_grl, file.path(mock_dir, "altius_tf_bindsites.rds"))

# 2. Mock GC frequencies
mock_gcfreq <- list(mock_motif_1 = matrix(0.25, nrow = 5, ncol = 4))
saveRDS(mock_gcfreq, file.path(mock_dir, "altius_motif_gcfreq.rds"))

# 3. Mock Genome GC
mock_gr <- GenomicRanges::GRanges("chr1:1-100")
mock_gr$GC_bias <- 0.5
mock_gr$GC_bin <- 1L
saveRDS(mock_gr, file.path(mock_dir, "genomewide_GC_hg38.rds"))

# Temporarily point the package to tempdir
old_opt <- getOption("methylTFRAnnotationHg38.datadir")
options(methylTFRAnnotationHg38.datadir = mock_dir)
```

# Usage

```{r usage}
library(methylTFRAnnotationHg38)

tf_bindsites <- getTFbindsites("altius")
gcfreqs <- getGCfreq("altius")
gc_dist <- getGenomeGC()
```

These are passed straight to `methylTFR::run_methyltfr()`.

# Available motif sets

```{r available_motifs, eval = FALSE}
methylTFRAnnotationHg38:::.MOTIF_SETS
```

Sets ending in `_distal` provide GC frequency tables computed
over distal regulatory regions only. They share the unrestricted
binding sites, so pass the base set name to `getTFbindsites()` 
and the `_distal` name to `getGCfreq()`.

# Session info

```{r session_info}
sessionInfo()
```

```{r cleanup, include = FALSE}
# HIDDEN CHUNK: Clean up the options
options(methylTFRAnnotationHg38.datadir = old_opt)
```
